---
title: "Winter Training 2026 Analysis"
format: revealjs
editor: visual
---

```{r}
#| warning: false
#| output: false
#| echo: false
## initial cleaning and wrangling done in excel
library(readr)
library(ggplot2)
library(tidyverse)
library(openintro)
library(ggeffects)
library(broom)
x500sData <- read_csv("500sData.csv")
x1000sData <- read_csv("1000sData.csv")

x500_long <- x500sData |> pivot_longer(cols = Interval1:Interval7, 
                                       names_to = "Interval", values_to = "Time") |>
  mutate("Time" = as.numeric(Time)) |>
  mutate("Time" = Time/60)

x500_line <- x500_long |> group_by(Trial, Interval) |>
  summarise(Mean = (mean(Time)))
  

x1000_long <- x1000sData |> pivot_longer(cols = Interval1:Interval5, 
                                       names_to = "Interval", values_to = "Time")|>
  mutate("Time" = as.numeric(Time)) |>
  mutate("Time" = Time/60)

x1000_line <- x1000_long |> group_by(Trial, Interval) |>
  summarise(Mean = (mean(Time)))
```

## Data Introduction
Training and practicing during the winter is proven to be difficult for rowing and most climates, especially the North Country, force rowers indoors for the season. The most common practice during the winter training period in rowing is the ergometer, commonly called an erg. The erg mimics the stroke that rowers take in the boat and instantly displays performance metrics to the rower. Metrics shown include the total time and distance of the workout, split per 500m (how long, in minutes and seconds, it takes the rower to complete 500m), and the stroke rate per minute.

## Data Introduction
The data was collected from the St. Lawrence University Women's Rowing Team during their 2026 winter training period. 

The women's rowing team ergs five times a week and records data on each piece that they complete including interval pieces and longer steady state pieces that build fitness and test pieces that show where their performance is at. This data includes 7x500m intervals, 5x1000m intervals, 2k test pieces, and 10k test pieces.

## Data Introduction
#### Variable Descriptions

| Variable | Description |
|--------------------|----------------------------------------------------|
| `ID` | Number indicating athlete |
| `Average` | Average time per 500m for each interval. |
| `IntervalX` | Average time for 500m for that interval. |
| `Rate` | Average strokes per minute during the piece (not included in 1000s). |
| `ClassYear` | What year the athlete is (Freshman, Sophomore, Junior, Senior) |

## Data Introduction

| Variable | Description |
|--------------------|----------------------------------------------------|
| `Interval` | Which interval (of 7 or 5) the time is for. |
| `Time` | Time (in seconds) for that interval. |
| `Trial` | Number indicating which number trial the piece is (first of the season, second, third). |
| `Mean` | Average time of all athletes per 500m for that interval and trial. |

## Questions of Interest

1.  How do athletes fare during many short interval pieces compared to fewer, longer intervals?

    -\> is there a similar trend between 7x500 and 5x1000?

    -\> what could be the explanation for these trends?

2.  What variables best predict 2k time?

    -\> and what does that prediction look like?

## 7x500m Intervals

```{r}
ggplot(data = x500_long, aes(x = Interval, y = Time)) +
  geom_point(aes(group = Trial, colour = factor(Trial)), alpha = 0.7) +
  geom_line(data = x500_line, aes(x = Interval, y = Mean, 
                                   group = Trial, colour = factor(Trial)), linewidth = 1.5) +
  scale_colour_manual(values = c("purple", "steelblue", "darkorange"), name = "Average of Trial:") +
  labs(y = "Seconds per 500 meters", colour = "AverageTrial", title = "7x500 meter Intervals",
       caption = "Data collected from St. Lawrence University Crew Team Winter 2026")+
  theme_minimal(base_size = 18)
```

## 5x1000m Intervals

```{r}
ggplot(data = x1000_long, aes(x = Interval, y = Time)) +
  geom_point(aes(group = Trial, colour = factor(Trial))) +
  geom_line(data = x1000_line, aes(x = Interval, y = Mean, 
                                   group = Trial, colour = factor(Trial)),linewidth = 1.5) +
  scale_colour_manual(values = c("purple", "darkorange"), name = "Average of Trial:") +
  labs(y = "Seconds per 500 meters", colour = "AverageTrial", title = "5x1000 meter Intervals",
       caption = "Data collected from St. Lawrence University Crew Team Winter 2026")+
  theme_minimal(base_size = 18)
```

## Linear Regression Visualization

```{r}
#| warning: false
#| output: false
#| echo: false
pred_10k_2k <- read_csv("pred_10k_2k.csv") |>
  mutate(first10k = as.numeric(first10k)) |>
  mutate(second10k = as.numeric(second10k)) |>
  mutate(k2 = as.numeric(k2)) |>
  mutate(k2 = k2/60) |>
  mutate(first10k = first10k/60) |>
  mutate(second10k = second10k/60) |>
  mutate(v1 = case_when(v1 == 0 ~ "No", v1 == 1 ~ "Yes"))

mod_2k <- lm(k2 ~ second10k + v1, data = pred_10k_2k)
mod_2k |> tidy()
glance(mod_2k)

ggpredict(mod_2k, terms = c("second10k", "v1"))

ggpredict(mod_2k, terms = c("second10k", "v1")) |> 
  as.data.frame(terms_to_colnames = TRUE) |>
  as_tibble()

pred_time <- ggpredict(mod_2k, terms = c("second10k",  "v1")) |>
  as.data.frame(terms_to_colnames = TRUE) |>
  mutate(v1 = recode(v1, `0` = "No", `1` = "Yes")) |>
  as_tibble()

pred_time <- pred_time |>
  arrange(v1, second10k) |>
  mutate(v1 = factor(v1, levels = c("No", "Yes")))
```

```{r}
ggplot(data = pred_10k_2k, aes(x = second10k, y = k2)) +
  geom_point(aes(colour = factor(v1)), alpha = 0.8) +
  geom_line(data = pred_time, aes(x = second10k, y = predicted, 
                                  group = interaction(v1), 
                                  colour = factor(v1)), linewidth = 1) +
  geom_ribbon(data = pred_time, aes(x = second10k, y = predicted,
      ymin = conf.low, ymax = conf.high, group = interaction(v1),
      fill = factor(v1)), alpha = 0.2) +
  labs(colour = "Varsity Boat", fill = "Varsity Boat", y = "2k time (in seconds)",
       caption = "Data collected from St. Lawrence University Crew Team Winter 2026",
       x = "10k time (in seconds)", title = "Predicting 2k time based on 10k time and Varsity Boat Status") +
  scale_colour_manual(values = c("steelblue", "darkorange")) +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  theme_minimal(base_size = 20)
```

